\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-07 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
update competing risk

We want to compare our hazard ratio estimates to that obtained from a
Cox regression. To obtain these estimates, we need to treat the
competing event as censoring:

\begin{Schunk}
\begin{Sinput}
library(survival)
# Treat competing event as censoring
model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
                   data = bmtcrr)
\end{Sinput}
\end{Schunk}

@@

\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.19)\\
Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.29, 0.93)\\
Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.48, 3.02)\\
Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.43, 6.5)\\
Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (2.09, 9.9)\\
\addlinespace
Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.54)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

@@

# Estimate absolute risk curve
risk_fg <- predict(model_fg, newdata, times = time_points)
\end{Sinput}
\end{Schunk}

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
\end{figure}
\end{Schunk}

Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both
case-base sampling and the Fine-Gray model. As we can see, the two
approaches agree quite well for AML; however, there seems to be a
difference of about 5\% between the two curves for ALL. This difference
does not appear to be significant: the curve from case-base sampling is
contained within a 95\% confidence band around the Fine-Gray absolute
risk curve (figure not shown).

@@

In this article, we presented the R package \pkg{casebase}, which
provides functions for fitting smooth parametric hazards and estimating
risk functions using case-base sampling. Our package also provides
several functions to produce graphical summaries of the data and the
results. We outlined the theoretical underpinnings of the approach, we
provided details about our implementation, and we illustrated the merits
of the approach and the package through three case studies.

@@

In the case studies above, we compared the performance of case-base
sampling with that of Cox regression and Fine-Gray models. In terms of
function interface, \pkg{casebase} uses a formula interface that is
closer to that of \code{glm}, in that the event variable is the only
variable appearing on the left-hand side of the formula. By contrast,
both \code{survival::coxph} and \code{timereg::comp.risk} use arrays
that capture both the event type and time. Both approaches to modeling
yield user-friendly code. However, in terms of output, both approaches
differ significantly. Case-base sampling produces smooth hazards and
smooth survival curves, whereas Cox regression and Fine-Gray models
produce stepwise CIFs and never explicitly model the hazard function.
Qualitatively, we showed that by using splines in the linear predictor,
all three models yielded similar curves. However, the smooth nature of
the output of \pkg{casebase} provides a more intuitive interpretation
for consumers of these predictions. In Table \ref{tab:compCBvsCox}, we
provide a side-by-side comparison between the Cox model and case-base
sampling.

\begin{widetable}
\caption{\label{tab:compCBvsCox}Comparison between the Cox model and case-base sampling}
\centering
\begin{tabular}[t]{llp{5cm}}
\toprule
Feature & Cox model & Case-base sampling\\
\toprule
Model type & Semi-parametric & Fully parametric\\\midrule
Time & Left hand side of the formula & Right hand side (allows flexible modeling of time)\\\midrule
Survival curve & Step function & Smooth-in-time curve\\\midrule
Non-proportional hazards & Interaction of covariates with time & Interaction of covariates with time\\\midrule
Model testing & Likelihood-ratio tests, score tests, & Use GLM framework\\
 & information criteria & (e.g.\ LRT, AIC, BIC)\\\midrule
\addlinespace
Competing risks & Difficult & Cause-specific CIFs\\
% Prediction & Kaplan-Meier-based & ROC, AUC, risk reclassification probabilities\\
\bottomrule
\end{tabular}
\end{widetable}

>>>>

We want to compare our hazard ratio estimates to that obtained from a
Cox regression.

\begin{Schunk}
\begin{Sinput}
library(survival)
# Prepare data for coxph
bmtcrr_cox <- transform(bmtcrr, 
                        id = seq_len(nrow(bmtcrr)),
                        Status = factor(Status))

model_cox <- coxph(Surv(ftime, Status) ~ Sex + D + Phase + Source + Age,
                   data = bmtcrr_cox, id = id)
\end{Sinput}
\end{Schunk}

@@

\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.74 & (0.43, 1.3) & 0.68 & (0.39, 1.21)\\
Disease & 0.52 & (0.29, 0.94) & 0.52 & (0.28, 0.94)\\
Phase (CR2 vs. CR1) & 1.20 & (0.48, 3) & 1.21 & (0.47, 3.09)\\
Phase (CR3 vs. CR1) & 1.59 & (0.41, 6.17) & 1.67 & (0.46, 6.08)\\
Phase (Relapse vs. CR1) & 4.09 & (1.9, 8.81) & 4.55 & (1.98, 10.46)\\
\addlinespace
Source & 1.70 & (0.55, 5.23) & 1.46 & (0.47, 4.53)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

@@

# Estimate CI curve
risk_fg <- predict(model_fg, newdata, times = time_points)
\end{Sinput}
\end{Schunk}

We can also estimate the CI for relapse using the Cox model and the
Aalen-Johansen estimator:

\begin{Schunk}
\begin{Sinput}
# Estimate absolute risk curve
risk_cox <- survfit(model_cox, newdata = newdata)
\end{Sinput}
\end{Schunk}

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./bmtcrr-risk-1} \caption{\label{fig:compAbsrisk} Cumulative Incidence curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Fine-Gray and Aalen-Johansen estimates.}\label{fig:bmtcrr-risk}
\end{figure}
\end{Schunk}

Figure \ref{fig:compAbsrisk} shows the cumulative incidence curves for
all three models. As we can see, all three approaches agree quite well
for AML; however, for ALL, there seems to be a difference of about 5\%
between the Fine-Gray curve and the curves estimated using case-base
sampling and Cox regression. This difference does not appear to be
significant: the curve from case-base sampling is contained within a
95\% confidence band around the Fine-Gray absolute risk curve (figure
not shown).

@@

In this article, we presented the R package \pkg{casebase}, which
provides functions for fitting smooth parametric hazards and estimating
risk functions using case-base sampling. Our package also provides
several functions to produce graphical summaries of the data and the
results. We outlined the theoretical underpinnings of the approach, we
provided details about our implementation, and we illustrated the merits
of the case-base framework and the package through three case studies.

@@

In the case studies above, we compared the performance of case-base
sampling with that of Cox regression and Fine-Gray models. In terms of
function interface, \pkg{casebase} uses a formula interface that is
closer to that of \code{glm}, in that the event variable is the only
variable appearing on the left-hand side of the formula. By contrast,
both \code{survival::coxph} and \code{timereg::comp.risk} use arrays
that capture both the event type and time. Both approaches to modeling
yield user-friendly code. However, in terms of output, both approaches
differ significantly. Case-base sampling produces smooth hazards and
smooth survival curves, whereas Cox regression and Fine-Gray models
produce stepwise CIFs and never explicitly model the hazard function.
Qualitatively, we showed that by using splines in the linear predictor,
all three models yielded similar curves. However, the smooth nature of
the output of \pkg{casebase} provides a more intuitive interpretation
for consumers of these predictions.
